Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-12-21 manuscript.tex
latest

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This dispersal Poisson distribution is denoted $p_i(x_i)$. Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').

@@

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we see that $j$ will invade if it is superior in any one of the three traits, but is otherwise identical to $i$. If the new genotype has the same competitive ability $c_j=c_i$, then $R_j\approx A_i$ and Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$. In the case that the new genotype has a different competitive ability, Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs. \eqref{eq:equil} and \eqref{eq:invad} respectively. Moreover, if $j$ invades in any of these cases, it will eventually exclude $i$, since it is strictly superior.

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{coex.pdf}
\caption{\label{fig:coex} Coexistence between $b$ ($c_i=1$, $b_i=1$) and $c$ ($c_j=2$, $b_j=0.7$) specialists, where $d_i=d_j=0.3$. Vertical axis shows frequency of the $c$-specialist predicted by Eq. \eqref{eq:master}.} 
\end{figure}

A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency. Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e. $A_i+R_i=A_j+R_j$). However, this coexistence scenario is not biologically relevant, since the tiniest deviation from $b_i/d_i=b_j/d_j$ will lead to the eventual exclusion of the genotype with greater $b_i/d_i$.

If the rare genotype $j$ arises due to mutation, then its initial low-density behavior is more complicated than the above invasion analysis suggests. The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance. During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (section ``Model''). However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq. \eqref{eq:master}. For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e. the mutant's fitness minus the population mean fitness). Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes. In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.

@@

For each genotype's dispersal, the counts of propagules across the unnocupied territories follows a multinomial distribution with equal probabilities of ``success'' for all territories. Thus, the $x_i$ in different territories are not independent random variables. However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$. Here we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.

>>>>

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$). Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).  Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').

@@

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we see that $j$ will invade if it is superior in any one of the three traits, but is otherwise identical to $i$. If the new genotype has the same competitive ability $c_j=c_i$, then $R_j\approx A_i$ and Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$. In the case that the new genotype has a different competitive ability but the same $b_i$ and $d_i$, Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs. \eqref{eq:equil} and \eqref{eq:invad} respectively. Moreover, if $j$ invades in any of these cases, it will eventually exclude $i$, since it is strictly superior.

@@

A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency. Mutual invasibility is not possible between $b$- and $d$-specialists (although genotypes with exactly $c_i=c_j$ and $b_i/d_i=b_j/d_j$ do not exclude each other --- this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e. $A_i+R_i=A_j+R_j$).

If the rare genotype $j$ arises due to mutation, then its initial low-density behavior is more complicated than the above invasion analysis suggests. The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance. During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (section ``Model''). However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq. \eqref{eq:master}. For mutants with fitness greater than the population mean fitness, this occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e. $s=\frac{\Delta n_i/n_i}{\sum_i\Delta n_i/n_i\times n_i/N}-1$). Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes. In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.

@@

For each genotype's dispersal, the counts of propagules across unnocupied territories follows a multinomial distribution with equal probabilities of landing in each territory. Thus, the $x_i$ in different territories are not independent random variables. However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$. Here we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.
